---
title: "'SciViews::R' - Statistical Models"
author: "Philippe Grosjean & Guyliann Engels"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{'SciViews::R' - Statistical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = 'center',
  fig.width = 7, fig.height = 5, out.width = "100%")
library(modelit)
library(tabularise)
library(equatiomatic)
library(chart)
library(dplyr)
```

The {modelit} package enhances statistical models calculated using `lm()`, `glm()` or `nls()` with nice charts and rich-formatted tables. It used both the `fun$type(data = ...., formula)` syntax. It also considers the labels of the variables in its outputs if a `label` attribute is defined.

## Example

Load the required packages for this example.

```{r}
library(modelit)
library(tabularise)
library(equatiomatic)
library(chart)
library(dplyr)
```

The `Loblolly` dataset is used to illustrate some {modelit} features. It focuses on the growth of Loblolly pine trees (*Pinus tadea* L. 1753). The dataset has three variables: `height`, `age`, and the `seed`.

```{r}
data("Loblolly", package = "datasets")
# convert feet to meters
Loblolly$height <- round(Loblolly$height * 0.3048, 2)
# Add labels and units
Loblolly <- svBase::labelise(Loblolly,
  label = list(height = "Heigth", age = "Age", Seed = "Seed"),
  units = list(height = "m", age = "years"))

# First and last rows of the dataset with tabularise
tabularise$headtail(Loblolly)
```

Here is a plot of the data with `chart()`, automatically using labels and units defined here above:

```{r}
chart(Loblolly, height ~ age %col=% Seed) +
  geom_line()
```

We only keep one measurement per tree (*alias* `Seed`) to get a dataset with independent observations to adjust a simple linear model.

```{r}
set.seed(3652)
pine <- dplyr::slice_sample(Loblolly, n = 1, by = Seed)
chart(pine, height ~ age) + 
  geom_point()
```

We fit a quadratic model to the data.

```{r}
pine_lm <- lm(data = pine, height ~ age + I(age^2))
summary(pine_lm)
```

Compare the "raw" summary of the **lm** object above with the version formatted using `tabularise()`.

```{r}
summary(pine_lm) |> 
  tabularise() 
```

The equation can be directly displayed in an R Markdown or Quarto document with the `equation(pine_lm)` code in an inline R chunk.

`r equation(pine_lm)`

Several plots are also available with `chart()`, starting with the graph showing the model fitted to the data:

```{r}
chart(pine_lm) # Equivalent to : chart$model(pine_lm)
```

The plots of the analysis of the residuals are `chart()` variants that can be indicated in a short form as `chart$<type>()` function and the possible types are: `"resfitted"`, `"qqplot"`, `"scalelocation"`, `"cooksd"`, `"resleverage"`, `"cookleverage"`, `"reshist"`, and `"resautocor"`. A composite plot, incorporating the four most used residual analysis charts, is obtained with the `"residuals"` type (note that the four charts are labeled A-D to facilitate their description in the legend).

```{r}
chart$residuals(pine_lm)
```


## Nonlinear model

Other objects are also supported by {modelit}. For instance, a nonlinear model can be fitted to the same dataset. A Gompertz model is more suitable for such growth measurements.

```{r}
pine_nls <- nls(data = pine, height ~ SSgompertz(age, a, b1, b2))
summary(pine_nls) |> 
  tabularise()
```

The plot of the model superposed to the data can be obtained once again with `chart()`.

```{r}
chart(pine_nls, lang = "fr")
```

